---
title: "Use of IlluminaHumanMethylationEPICv2manifest"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use of IlluminaHumanMethylationEPICv2manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```

This package provides manifest for Illumina methylation EPIC array v2.0. The data is based on the file https://support.illumina.com/content/dam/illumina-support/documents/downloads/productfiles/methylationepic/MethylationEPIC%20v2%20Files.zip from 
  https://support.illumina.com/array/array_kits/infinium-methylationepic-beadchip-kit/downloads.html.

A script for creating the data object in this package is contained in
  `system.files("scripts", "manifest.R", package = "IlluminaHumanMethylationEPICv2manifest")`, 
which is slightly adjusted from the original script in the **IlluminaHumanMethylationEPICmanifest** package.

When using with the **minfi** package, you can manually set the `"array"` element by providing the prefix (removing the string "manifest"):

```{r, eval = FALSE}
RGset = read.metharray.exp(...)

annotation(RGset)["array"] = "IlluminaHumanMethylationEPICv2"

# explained in the IlluminaHumanMethylationEPICv2anno.20a1.hg38 package
annotation(RGset)["annotation"] = "20a1.hg38"
```

## Compare EPIC array v1 and v2 probes

The following code compares probes in **IlluminaHumanMethylationEPICv2manifest** and **IlluminaHumanMethylationEPICmanifest**.

```{r}
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylationEPICv2manifest)

probe1 = getManifestInfo(IlluminaHumanMethylationEPICmanifest, "locusNames")
probe2 = getManifestInfo(IlluminaHumanMethylationEPICv2manifest, "locusNames")

# probes can be duplicated
probe1 = unique(probe1)
probe2 = unique(probe2)
```

Number of probes:

```{r}
length(probe1)
length(probe2)
```

Overlap of probes:

```{r, fig.width = 6, fig.height = 6}
library(eulerr)
plot(euler(list(v1 = probe1, v2 = probe2)), quantities = TRUE)
```

## What if using the wrong array version?

We get the demo dataset of EPIC array v2.0 from the following link: https://support.illumina.com/array/array_kits/infinium-methylationepic-beadchip-kit/downloads.html, with the name "Infinium MethylationEPIC v2.0 Demo Data Set (iScan)".


We randomly picked one pair of files. To create this vignette dynamically, we hosted the files on GitHub and now we download
them and put them in a temporary folder. `206891110001_R01C01.zip` contains one `*_Grn.idat` and corresponding `*_Red.idat` file.

```{r}
tempdir = tempdir()
datadir = paste0(tempdir, "/206891110001")
dir.create(datadir, showWarnings = FALSE)

url = "https://github.com/jokergoo/IlluminaHumanMethylationEPICv2manifest/files/11008723/206891110001_R01C01.zip"
local = paste0(tempdir, "/206891110001_R01C01.zip")
download.file(url, dest = local, quiet = TRUE)

unzip(local, exdir = datadir)
```

Next we use `minfi::read.metharray.exp()` to import the original data.

```{r}
library(minfi)
RGset = read.metharray.exp(datadir)
RGset
```

We can see there are more than 1 million probes. Note **minfi** cannot
automatically detect the array type of this data ("array" is assigned to
"Unknown").

If we manually assign the EPIC array v1.0 to `RGset` and perform preprocessing:

```{r}
annotation(RGset)["array"] = "IlluminaHumanMethylationEPIC"
preprocessRaw(RGset)
```

Only around 100K probes remain, which is definitely wrong.

While if we choose the correct v2.0 array type:

```{r}
annotation(RGset)["array"] = "IlluminaHumanMethylationEPICv2"
preprocessRaw(RGset)
```

More than 900K probes remain, which fits the design of the EPIC array v2.0.

```{r, echo = FALSE}
unlink(datadir)
unlink(local)
```

## Session info

```{r}
sessionInfo()
```
